### 2016-11-04 abstract.tex
updates

\begin{abstract}
How much can pruning algorithms teach us about the fundamentals of learning representations in neural networks? A lot, it turns out. Neural network model compression has become a topic of great interest in recent years, and many different techniques have been proposed to address this problem. In general, this is motivated by the idea that smaller models typically lead to better generalization. At the same time, the decision of what to prune and when to prune it necessarily forces us to confront our assumptions about how neural networks actually learn to represent patterns in data. In this work we set out to test several long-held hypotheses about neural network learning representations and numerical approaches to pruning. To accomplish this we first reviewed the historical literature and derived a novel algorithm to prune whole neurons from optimally trained networks using a second-order Taylor method. We then set about testing the performance of our algorithm and analyzing the quality of the decisions it made. As a baseline for comparison we used a first-order Taylor method based on the Skeletonization algorithm and an exhaustive brute-force serial pruning algorithm. Our proposed algorithm worked well compared to a first-order method, but not nearly as well as the brute-force method. Our error analysis led us to question the validity of many widely-held assumptions behind pruning algorithms in general, and the trade-offs we often make in the interest of reducing computational complexity. We discovered that there is a straightforward way, however expensive, to serially prune 40-60\% of the neurons in a trained network with minimal effect on the learning representation and without any re-training. 
\end{abstract}

>>>>

\begin{abstract}
How much can pruning algorithms teach us about the fundamentals of learning representations in neural networks? A lot, it turns out. Neural network model compression has become a topic of great interest in recent years, and many different techniques have been proposed to address this problem. In general, this is motivated by the idea that smaller models typically lead to better generalization. At the same time, the decision of what to prune and when to prune it necessarily forces us to confront our assumptions about how neural networks actually learn to represent patterns in data. In this work we set out to test several long-held hypotheses about neural network learning representations and numerical approaches to pruning. To accomplish this we first reviewed the historical literature and derived a novel algorithm to prune whole neurons from optimally trained networks using a second-order Taylor method. We then set about testing the performance of our algorithm and analyzing the quality of the decisions it made. As a baseline for comparison we used a first-order Taylor method based on the Skeletonization algorithm and an exhaustive brute-force serial pruning algorithm. Our proposed algorithm worked well compared to a first-order method, but not nearly as well as the brute-force method. Our error analysis led us to question the validity of many widely-held assumptions behind pruning algorithms in general and the trade-offs we often make in the interest of reducing computational complexity. We discovered that there is a straightforward way, however expensive, to serially prune 40-60\% of the neurons in a trained network with minimal effect on the learning representation and without any re-training. 
\end{abstract}

### 2016-11-04 introduction.tex
updates

\section{Introduction}\label{sec1}
%Neural network pruning algorithms were first popularized by \cite{sietsma1988neural} as a mechanism to determine the proper size of a network required to solve a particular problem. To this day, network design and optimal pruning remain inherently difficult tasks. For problems which cannot be solved using linear threshold units alone, \cite{baum1989size} demonstrate there is no way to precisely determine the appropriate size of a neural network a priori given any random set of training instances. %Using too few neurons inhibits learning, and so in practice it is common to attempt to over-parameterize networks initially using a large number of hidden units and weights. However, as \cite{chauvin1990generalization} writes, this approach can lead to over-fitting as the network's unnecessary free parameters start to latch on to idiosyncrasies in the training data.

Pruning algorithms, as comprehensively surveyed by \cite{reed1993pruning}, are a useful set of heuristics designed to identify and remove network parameters which do not contribute significantly to the output of the network and potentially inhibit generalization performance. At the time of Reed's writing, reducing network size was also a practical concern, as smaller networks are preferable in situations where computational resources are scarce. In this paper we are particularly concerned with application domains where space is limited and network size constraints must be imposed with minimal impact on performance.

>>>>

\section{Introduction \& Literature Review}\label{sec1}
Pruning algorithms, as comprehensively surveyed by \cite{reed1993pruning}, are a useful set of heuristics designed to identify and remove network elements which do not contribute significantly to the output of the network and potentially inhibit generalization performance. This is motivated by the well-known tendency of neural networks to over-fit to the idiosyncrasies of their training data when given too many parameters or too few input patterns from which to generalize.

Neural network pruning algorithms were first popularized by \cite{sietsma1988neural} as a mechanism to determine the proper size of a network required to solve a particular problem. To this day, network design and optimal pruning remain inherently difficult tasks. For problems which cannot be solved using linear threshold units alone, \cite{baum1989size} demonstrate there is no way to precisely determine the appropriate size of a neural network a priori given any random set of training instances. Using too few neurons inhibits learning, and so in practice it is common to attempt to over-parameterize networks initially using a large number of hidden units and weights. However, as \cite{chauvin1990generalization} writes, this approach can lead to over-fitting as the network's unnecessary free parameters start to latch on to idiosyncrasies in the training data.

%\section{Related Work}

The generalization performance of neural networks has been well studied, and apart from pruning algorithms many heuristics have been used to avoid overfitting, such as dropout (\cite{srivastava2014dropout}), maxout (\cite{goodfellow2013maxout}), and cascade correlation (\cite{fahlman1989cascade}), among others. However, these algorithms do not explicitly prioritize the reduction of network memory footprint as a part of their optimization criteria per se, (although in the opinion of the authors, Fahlman's cascade correlation architecture holds great promise in this regard.) Computer memory size and processing capabilities have improved so much since the introduction of pruning algorithms in the late 1980s that space complexity has become a relatively negligible concern. The proliferation of cloud-based computing services has furthermore enabled mobile and embedded devices to leverage the power of massive data and computing centers remotely. In this domain, however, it is also reasonable to suggest that certain performance-critical applications running on low-resource devices could benefit from the ability to use neural networks locally.

At present there are few (if any) mechanisms specifically designed to shrink neural networks down in order to meet an externally imposed constraint on byte-size in memory. Without explicitly removing parameters from the network, one could use weight quantization to reduce the number of bytes used to represent each weight parameter, as investigated by \cite{balzer1991weight}, \cite{dundar1994effects}, and \cite{hoehfeld1992learning}. Of course, this method can only reduce the size of the network by a factor proportional to the byte-size reduction of each weight parameter.

Recently, some other network compression methods have been proposed. One such method which has recently gained popularity (\cite{prabhavalkar2016svd}) uses the singular values of a trained weight matrix as basis vectors from which to derive a compressed hidden layer. Some other recent works like \cite{Anders2016quant} have tried successfully to achieve compression through weight quantization followed by encoding while some others like \cite{deepcompression2016} have tried to build on top of that by adding weight-pruning as a preceding step to quantization and encoding.

If we wanted to continually shrink a network to its absolute minimal size in an optimal manner, we might accomplish this using any number of off-the-shelf pruning algorithms, such as Skeletonization (\cite{mozer1989skeletonization}), Optimal Brain Damage (\cite{lecun1989optimal}), or later variants such as Optimal Brain Surgeon (\cite{hassibi1993second}). In fact, we borrow much of our inspiration from these antecedent algorithms, with one major variation:\textit{ Instead of pruning individual weights, we prune entire neurons, thereby eliminating all of their incoming and outgoing weight parameters in one go, resulting in more memory saved, faster.}

Scoring and ranking individual weight parameters in a large network is computationally expensive, and generally speaking the removal of a single weight from a large network is a drop in the bucket in terms of reducing a network's core memory footprint.  We argue that pruning neurons instead of weights is more efficient computationally as well as practically in terms of quickly reaching a target reduction in memory size. Our approach also attacks the angle of giving downstream applications a realistic expectation of the minimal increase in error resulting from the removal of a specified percentage of neurons. Such trade-offs are unavoidable, but performance impacts can be limited if a principled approach is used to find candidate neurons for removal.

Too many free parameters in a Neural Network lead to overfitting. Regardless of the number of weights used in a given network, as \cite{segee1991fault} assert, the representation of a learned function approximation is almost never evenly distributed over the hidden units, and the removal of any single hidden unit at random can actually result in a total network fault. \cite{mozer1989using} suggest that only a subset of the hidden units in a neural network actually latch on to the invariant or generalizing properties of the training inputs, and the rest learn to either mutually cancel each other out or begin over-fitting to the noise in the data. We leverage this idea in the current work to rank all neurons in pre-trained networks based on their effective contributions to the overall performance. We then remove the unnecessary neurons to reduce the network's footprint. Through our experiments we not only concretely validate the theory put forth by \cite{mozer1989using} but we also successfully build on  it to prune networks to 40 to 60 \% of their original size without any major loss in performance.
